{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-08-21 ms.tex

When modelling a patch, the focus is on changes to the size-
density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
individuals with traits \(x\) and height \(h\) within a patch of age \(a\).
Density is measured as individuals per unit height and per unit ground area.
Modelling the development of  \(n(x,h,a)\) requires that both the initial size
distribution and the inflow of new recruits be specified. We have primarily
been interested in dynamics of a patch recovering from disturbance and have
therefore started with an empty patch and constant flow of seeds from a global
seed rain (Fig. \ref{fig:schematic}b).

Our approach for modelling size-structured population dynamics is
based on the Escalator Boxcar Train (\textsc{ebt}) technique
\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves
the \textsc{pde} describing development of \(n\) by approximating the
density function with a collection of cohorts spanning the size
spectrum. Following a disturbance, a series of cohorts are introduced
into each patch. These cohorts are then transported along the
characteristic equations of the \textsc{pde}. Biologically these are
the growth trajectories of individuals, conditioned on them surviving
(e.g. Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along the trajectories
(Fig. \ref{fig:patch}b).

We introduce two extensions to the \textsc{ebt}: i) a new method for
estimating the size-density distribution and ii) a technique for
handling strong size-asymmetric competitive feedbacks, such as occur
under strong competition for light (see Appendix \ref{sec:demography}
for details). The original \textsc{ebt} proceeds by approximating the
first and second moments of the density distribution \(n\) within each
cohort \citep{Deroos-1997}.
%
In {\plant} we use a different approach. Instead of tracking first and
second moments of the size-distribution within cohorts, we track a
point mass estimate of \(n\) along the trajectory for the boundary of
each cohort. We found this approach preferable because it does not
require us to maintain a separate set of equations for the smallest
cohort.

@@

In {\plant} we assume an infinite number of patches, all experiencing
the same disturbance regime and sharing a common seed dispersal pool
(i.e., the island model; Fig. \ref{fig:schematic}b). We also assume
that the disturbances remove all established vegetation.  With this
approach we can scale up from simulating a single patch to simulate an
entire metapopulation.  The frequency-density \(p(a)\) of patches aged
\(a\) in the landscape is needed. Assuming there are a large number of
patches means the dynamics of \(p\) behave deterministically and can
be approximated via a second \textsc{pde} (see Appendix
\ref{sec:demography} for details). Scaling from patch to
metapopulation is then achieved by weighting the temporal dynamics of
a single patch by \(p(a)\), the relative abundance of patches aged
\(a\) in the metapopulation (Fig. \ref{fig:patch}c).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
\caption{Population dynamics for two species competing within a patch.
  Red and blue lines are low and high \textsc{lma} species from
  Fig. \ref{fig:plant}.  Lines in panel (a) show growth trajectories
  for individuals on the boundary of successive ``cohorts'' for each
  species, using a schedule of introduction times generated by
  {\plant}'s adaptive algorithm.  The initial growth rate advantage of
  the low \textsc{lma} species (see Figure \ref{fig:plant}c) means it
  quickly over tops the high \textsc{lma} species, suppressing its
  growth.  Self-thinning creates space for the high \textsc{lma}
  species to establish below the canopy of the faster species. Panel
  (b) shows the same trajectories as in panel (a), but with shading
  indicating the density $n$ of individuals at given size and patch
  age (light = low density, dark = high density).  (c) The leaf area
  index at ground level for each species.  The dotted line is the
  total leaf area index (summed across both species), and the lines on
  the right side are average when integrated over the whole
  metapopulation structure.  See Appendix \ref{sec:patch} for further
  details and code.}
\label{fig:patch}
\end{figure}

>>>>

When modelling a patch, the focus is on changes to the size-
density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density of
individuals with traits \(x\) and height \(h\) within a patch of age \(a\).
Density is measured as individuals per unit height and per unit ground area.
Modelling the development of  \(N(x, H,a)\) requires that both the initial size
distribution and the inflow of new recruits be specified. We have primarily
been interested in dynamics of a patch recovering from disturbance and have
therefore started with an empty patch and constant flow of seeds from a global
seed rain (Fig. \ref{fig:schematic}b).

Our approach for modelling size-structured population dynamics is
based on the Escalator Boxcar Train (\textsc{ebt}) technique
\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves
the \textsc{pde} describing development of \(N\) by approximating the
density function with a collection of cohorts spanning the size
spectrum. Following a disturbance, a series of cohorts are introduced
into each patch. These cohorts are then transported along the
characteristic equations of the \textsc{pde}. Biologically these are
the growth trajectories of individuals, conditioned on them surviving
(e.g. Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along the trajectories
(Fig. \ref{fig:patch}b).

We introduce two extensions to the \textsc{ebt}: i) a new method for
estimating the size-density distribution and ii) a technique for
handling strong size-asymmetric competitive feedbacks, such as occur
under strong competition for light (see Appendix \ref{sec:demography}
for details). The original \textsc{ebt} proceeds by approximating the
first and second moments of the density distribution \(N\) within each
cohort \citep{Deroos-1997}.
%
In {\plant} we use a different approach. Instead of tracking first and
second moments of the size-distribution within cohorts, we track a
point mass estimate of \(N\) along the trajectory for the boundary of
each cohort. We found this approach preferable because it does not
require us to maintain a separate set of equations for the smallest
cohort.

@@

In {\plant} we assume an infinite number of patches, all experiencing
the same disturbance regime and sharing a common seed dispersal pool
(i.e., the island model; Fig. \ref{fig:schematic}b). We also assume
that the disturbances remove all established vegetation.  With this
approach we can scale up from simulating a single patch to simulate an
entire metapopulation.  The frequency-density \(P(a)\) of patches aged
\(a\) in the landscape is needed. Assuming there are a large number of
patches means the dynamics of \(P\) behave deterministically and can
be approximated via a second \textsc{pde} (see Appendix
\ref{sec:demography} for details). Scaling from patch to
metapopulation is then achieved by weighting the temporal dynamics of
a single patch by \(P(a)\), the relative abundance of patches aged
\(a\) in the metapopulation (Fig. \ref{fig:patch}c).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
\caption{Population dynamics for two species competing within a patch.
  Red and blue lines are low and high \textsc{lma} species from
  Fig. \ref{fig:plant}.  Lines in panel (a) show growth trajectories
  for individuals on the boundary of successive ``cohorts'' for each
  species, using a schedule of introduction times generated by
  {\plant}'s adaptive algorithm.  The initial growth rate advantage of
  the low \textsc{lma} species (see Figure \ref{fig:plant}c) means it
  quickly over tops the high \textsc{lma} species, suppressing its
  growth.  Self-thinning creates space for the high \textsc{lma}
  species to establish below the canopy of the faster species. Panel
  (b) shows the same trajectories as in panel (a), but with shading
  indicating the density $N$ of individuals at given size and patch
  age (light = low density, dark = high density).  (c) The leaf area
  index at ground level for each species.  The dotted line is the
  total leaf area index (summed across both species), and the lines on
  the right side are average when integrated over the whole
  metapopulation structure.  See Appendix \ref{sec:patch} for further
  details and code.}
\label{fig:patch}
\end{figure}
